Tutorial: fitting mixed mixture models on text writing data

### 2023-02-07 tutorial/tutorial.Rmd
Swap prob in stan model

The model returns population estimates for the average typing speed of fluent keystroke transitions ($\beta$), the slowdown for hesitant transitions ($\delta$) and the frequency of hesitant transitions ($\theta$) for each level of a categorical predictor after taking into account three sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor; (3) keystroke intervals for some bigrams are slower than for others.

$$

$$

@@

The Stan model can be loaded into R using the `stan_model` function and assign it to the object `molg`. The "molg.stan" file needs to be available in the directory "stan" located inside the current working directory.

@@

```{r}
# Parameters to keep in output
pars <- c("beta",                        # fluent typing
          "delta",                       # disfluency slowdown
          "theta",                       # mixing proportion on logit scale
          "prob",                        # mixing proportion as probability
          "theta_s",                     # by-participant disfluency probability (logit scale)
          "prob_s",                      # by-participant disfluency probability
          "sigma",                       # variance component
          "sigma_diff", "sigmap_e", "sigma_e",  # variance by mixture component
          "sigma_u",                     # variance for random ppt 
          "log_lik",                     # log likelihood (for model comparison) 
          "y_tilde")                     # predicted data 
```

@@

```{r eval = F}
# Fit model
fit_molg <- sampling(molg, 
                     data = data_list,
                     iter = iterations,
                     warmup = warmup,
                     chains = nchains, 
                     cores = ncores,
                     pars = pars, # Parameters to keep.
                     save_warmup = FALSE, # Don't save the warmup samples.
                     seed = 365, 
                     control = list(adapt_delta = 0.96,  # default: 0.9
                                    max_treedepth = 14)) # default: 10
```

@@

```{r}
print(fit_molg, pars = c("beta", "delta", "theta"), probs = c(.025, .975))
```

@@

@@

Model convergence can be established in two practically simple techniques. First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain. We will look at the population-level parameters (names are determined by the model) corresponding to the Greek symbols used **in the manuscript** and assign those to `pars`.

```{r}
pars <- c("beta", "delta", "theta", "sigma", "sigmap_e", "sigma_e")
```

@@

```{r}
# Check convergence
stan_trace(m, pars = pars, alpha = .5)
```

@@

```{r}
summary(m, pars=pars)$summary %>% as.data.frame() %>% 
  rownames_to_column("parameter") %>% 
  select(parameter, Rhat)
```

@@

The model used the posterior parameter values to simulate hypothetical data sets. This happened for every iteration for every chains.

```{r echo = F, include=F}
(total_samples <- (iterations - warmup) * n_chains)
```

@@

```{r}
y_tilde <- as.matrix(m, pars = "y_tilde")          # extract simulated data sets
N <- 50                                            # number of simulations to use
total_samples <- (iterations - warmup) * n_chains  # total number of samples
rnd_sims <- sample(total_samples, N)               # created random indices
y_tilde_sample <- y_tilde[rnd_sims,]               # draw N random simulations 
```

@@

```{r }
(total_samples <- (iterations - warmup) * n_chains)
```

@@

The `names` function can be used to remind us of the parameter names used in the model (we reduced the output to the first 15 parameter names). The meaning of the parameters is described **in the manuscript** and can be obtained from the Stan code. Indices refer to the copy-task component (1 = consonants; 2 = LF bigrams) and to the participant identifier for by-participant parameters (indicated by `_s`).

```{r}
names(m)[1:15] 
```

@@

```{r}
stan_hist(m, pars = pars)
```

@@

```{r}
print(m, pars = pars, probs = c(.025,.975))
```

@@

The `plot` function shows the posterior probability distribution of the three parameters for the consonants task (indicated as 1) and the LF-bigrams task (indicated as 2) summarised as median and 95% PI.

```{r}
plot(m, pars = pars, ci_level = .95) # ci = credible interval
```

@@

```{r}
posterior <- as.data.frame(m, pars) %>% as_tibble() 
posterior
```

@@

```{r}
posterior_long <- pivot_longer(posterior, everything(), 
             names_to = c(".value", "component"), 
             names_pattern = "(.*)\\[(.)\\]")

posterior_long
```

This code is then transforming `beta` and `delta` to msecs using the exponential function `exp` to un-log the values. In order to transform the slowdown `delta` into msecs we need to add `beta` before using the exponential function; we can then subtract beta again. The `recode` function changes the component indices from 1 to "Consonants" and 2 to "LF bigrams".

```{r}
posterior_in_msecs <- mutate(posterior_long,
                             delta = exp(beta + delta) - exp(beta),
                             beta = exp(beta),
                             component = recode(component, `1` = "Consonants",
                                                           `2` = "LF bigrams"))
posterior_in_msecs
```

@@

```{r}
pivot_longer(posterior_in_msecs, beta:theta, names_to = "parameter") %>%
  ggplot(aes(x = value, colour = component, fill = component)) +
  geom_histogram(position = "identity", alpha = .25) +
  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
  scale_fill_brewer("Copy-task\ncomponents", palette = "Dark2") +
  scale_color_brewer("Copy-task\ncomponents", palette = "Dark2") +
  labs(x = "Posterior parameter estimate")
```

>>>>

The model returns population estimates for the average typing speed of fluent keystroke transitions ($\beta$), the slowdown for hesitant transitions ($\delta$) and the frequency of hesitant transitions ($\theta$) for each level of a categorical predictor after taking into account two sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.

$$
\text{iki}_{i} \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_i, \sigma_{e'_\text{location[i]}}^2) +\\
  (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_i, \sigma_{e_\text{location[i]}}^2)\\
		\text{where}\\
		\delta \sim \text{N}(0,1)\\
		\text{constraint: } \delta > 0
$$

@@

The Stan model can be loaded and translated into an S4 class object using the `stan_model` function and assign it to the object `molg`. The "molg.stan" file needs to be available in the directory "stan" located inside the current working directory.

@@

```{r}
# Parameters to keep in output
pars <- c("beta",                        # fluent typing
          "delta",                       # disfluency slowdown
          "theta",                       # mixing proportion on logit scale
          "prob",                        # mixing proportion as probability
          "prob_s",                      # by-participant disfluency probability
          "sigma",                       # variance component
          "sigma_diff", "sigmap_e", "sigma_e",  # variance by mixture component
          "sigma_u",                     # variance for random ppt 
          "log_lik",                     # log likelihood (for model comparison) 
          "y_tilde")                     # predicted data 
```

@@

```{r eval = F}
# Fit model
fit_molg <- sampling(molg, 
                     data = data_list,
                     iter = iterations,
                     warmup = warmup,
                     chains = nchains, 
                     cores = ncores,
                     pars = pars, # Parameters to keep.
                     save_warmup = FALSE, # Don't save the warmup samples.
                     seed = 365, 
                     # Improve sampling efficiency
                     control = list(adapt_delta = 0.96,  # default: 0.9
                                    max_treedepth = 14)) # default: 10
```

@@

@@

Inferred posterior estimates shown as average and 95% probability interval can be extracted using the `print` function for each transition location and the three parameter values of interest.

```{r}
print(fit_molg, pars = c("beta", "delta", "prob"), probs = c(.025, .975))
```

## Model comparsion

Similar to above we will compare the three models -- the mixture of log-Gaussians, and the uni-model Gaussian and log-Gaussian models -- using leave-one-out cross-validation. Because the mixture model was fitted directly in Stan and the uni-model models were fitted in `brms`, we need to extract the fit statistics before comparing the models using functions provided by the package `loo`.

```{r}
library(loo)
# Mixture of log-Gaussians
log_lik <- extract_log_lik(fit_molg, merge_chains = F) 
r_eff <- relative_eff(exp(log_lik)) # relative effective sample size
loo_molg <- loo(log_lik, r_eff = r_eff)

# Uni-modal log-Gaussian
loo_lgaus <- loo(fit_lgaus)

# Uni-modal Gaussian
loo_gaus <- loo(fit_gaus)
```

As described above we used leave-one-out cross-validation to compare the mixture of log-Gaussians model to each the uni-modal Gaussian and log-Gaussian model. Their difference is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`. Models are ordered from the model with the highest predictive performance to the lowest (i.e. `elpd_loo`).

```{r}
loo_comp <- loo_compare(loo_gaus, loo_lgaus, loo_molg)
```

To extract the model comparisons and fit statistics:

```{r}
loo_comp %>% 
  as.data.frame() %>% 
  rownames_to_column("models") %>% 
  mutate(across(models, recode, model3 = "fit_molg"),
         across(where(is.numeric), round, 1))  
```

The model comparison shows that the model `fit_molg` (the mixture of log-Gaussian ) is approximately `r round(abs(loo_comp[2,1] / loo_comp[2,2]),0)` times better than `fit_lgaus` (try `abs(loo_comp[2,1] / loo_comp[2,2])`).

@@

Model convergence can be established in two practically simple techniques. First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain. We will look at the population-level parameters (names are determined in the Stan code) and assign those to `pars`.

@@

```{r}
# Check convergence
stan_trace(fit_molg, pars = pars, alpha = .5)
```

@@

```{r}
summary(fit_molg, pars=pars)$summary %>% 
  as.data.frame() %>% 
  rownames_to_column("parameter") %>% 
  select(parameter, Rhat)
```

@@

```{r echo = F, include = F}
(total_samples <- (iterations - warmup) * nchains)
```

The model used the posterior parameter values to simulate hypothetical data sets. This happened for every iteration for every chains. In other words there is a total of `r scales::comma(total_samples)` simulated data sets (and samples per parameter) available to us because

```{r}
(total_samples <- (iterations - warmup) * nchains)
```

@@

```{r}
y_tilde <- as.matrix(fit_molg, pars = "y_tilde")   # extract simulated data sets
N <- 50                                            # number of simulations to use
rnd_sims <- sample(total_samples, N)               # created random indices
y_tilde_sample <- y_tilde[rnd_sims,]               # draw N random simulations 
```

@@

@@

The `names` function can be used to remind us of the parameter names used in the model (we reduced the output to the first 15 parameter names). The meaning of the parameters is described above and is determined in the Stan code. Indices refer to the transition locations (1 = before-sentence; 2 = before-word; 3 = within-word) and to the participant identifier for by-participant parameters (indicated by `_s`).

```{r}
names(fit_molg)[!str_detect(names(fit_molg), "log_lik|y_tilde")]
```

@@

```{r}
stan_hist(fit_molg, pars = pars)
```

@@

```{r}
print(fit_molg, pars = pars, probs = c(.025,.975))
```

@@

The `plot` function shows the posterior probability distribution of the three parameters for before-sentence transitions (indicated as 1), before-word transitions (indicated as 2), and within-word transitions summarised as median and 95% PI.

```{r}
plot(fit_molg, pars = pars, ci_level = .95) # ci = credible interval
```

@@

```{r}
posterior <- as.data.frame(fit_molg, pars) %>% as_tibble() 
posterior
```

@@

```{r}
posterior_long <- pivot_longer(posterior, everything(), 
             names_to = c(".value", "location"),
             names_pattern = "(.*)\\[(.)\\]") 
posterior_long
```

This code is then transforming `beta` and `delta` to msecs using the exponential function `exp` to un-log the values. In order to transform the slowdown `delta` into msecs we need to add `beta` before using the exponential function; we can then subtract beta again. `theta` samples are currently on the logit scale and can be transformed to probabilities using `brms::inv_logit_scaled`. The `recode` function changes the component indices from 1 to "before-sentence", 2 to "before-word", 3 to "within-word".

```{r}
posterior_in_msecs <- mutate(posterior_long,
                             delta = exp(beta + delta) - exp(beta),
                             across(beta, exp),
                             across(theta, inv_logit_scaled),
                             across(location, recode, `1` = "before-sentence",
                                                      `2` = "before-word",
                                                      `3` = "within-word"))
posterior_in_msecs
```

@@

```{r}
pivot_longer(posterior_in_msecs, beta:theta, names_to = "parameter") %>%
  ggplot(aes(x = value, colour = location, fill = location)) +
  geom_histogram(position = "identity", alpha = .25) +
  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
  scale_fill_brewer("Transition\nlocation", palette = "Dark2") +
  scale_color_brewer("Transition\nlocation", palette = "Dark2") +
  labs(x = "Posterior parameter estimate")
```
